---
title: "Novelty, Transience and Resonance in Human-AI Stories"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
---

## Summary

- Humans produce higher novelty (surprisal) and higher transience than the model, indicating stronger deviations from earlier story context.
- Both agents display a positive novelty–resonance relationship (innovation bias), but the slope is steeper for humans.
- This indicates that surprising human turns exert greater forward influence on narrative development.
- AI contributions show lower variance and lower informational impact relative to human inputs.
- Humans are the primary drivers of narrative innovation and thematic redirection.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

pacman::p_load(tidyverse, lme4, lmerTest, here, emmeans)

theme_set(theme_minimal(base_size = 14))
GRAPHS_DIR <- here("analysis", "figures")
dir.create(GRAPHS_DIR, showWarnings = FALSE, recursive = TRUE)

EXCLUDED_CONVERSATIONS <- c(
  "conv_0bb56093-3033-4615-bb70-ebfa4135589a",
  "conv_0f18b30f-7d4b-4681-b98e-a0ff4f2b5256",
  "conv_72218cb5-e59c-4c93-a4b9-a057fe5dad80",
  "conv_7c23347e-6172-4c84-9fd0-45ef34290bd5",
  "conv_ed575a06c11d42358e3eeb7826d2f959",
  "conv_a63a08273d0a4704a7638e4cd6850225"
)
```

# 1. Data Loading and Preprocessing

## 1.1 Load Data

```{r load-data}
df_res_field <- read_csv(here("data/human-ai/processed/novelty_scores.csv")) %>%
  rename(
    user = author_1,
    ai = author_2,
    user_surprise = author_1_surprise,
    ai_surprise = author_2_surprise,
    user_transience = author_1_transience,
    ai_transience = author_2_transience
  ) %>%
  filter(!is.na(user_surprise) & !is.na(ai_surprise)) %>%
  mutate(
    user_resonance = user_surprise - user_transience,
    ai_resonance = ai_surprise - ai_transience,
    datatype = "Field"
  ) %>%
  drop_na(user, ai) %>%
  filter(!conversation_id %in% EXCLUDED_CONVERSATIONS, turn > 1)
```

## 1.2 Reshape Data

```{r reshape-data}
df_res_field <- df_res_field %>%
  mutate(
    user_nwords = str_count(user, "\\S+"),
    ai_nwords = str_count(ai, "\\S+")
  )

df_res_long <- df_res_field %>%
  pivot_longer(
    cols = c(user_surprise, user_transience, user_nwords,
             ai_surprise, ai_transience, ai_nwords),
    names_to = c("agent", "metric"),
    names_pattern = "(user|ai)_(surprise|transience|nwords)"
  ) %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  mutate(
    resonance = surprise - transience,
    agent = factor(agent, levels = c("user", "ai"), labels = c("User", "LLM"))
  )
```

## 1.3 Descriptive Statistics

```{r descriptive-stats}
desc_stats <- df_res_long %>%
  group_by(agent) %>%
  summarise(
    n = n(),
    mean_surprise = mean(surprise, na.rm = TRUE),
    sd_surprise = sd(surprise, na.rm = TRUE),
    mean_transience = mean(transience, na.rm = TRUE),
    sd_transience = sd(transience, na.rm = TRUE),
    mean_resonance = mean(resonance, na.rm = TRUE),
    sd_resonance = sd(resonance, na.rm = TRUE)
  )

print(desc_stats)

knitr::kable(desc_stats, caption = "Descriptive Statistics for Novelty Metrics by Agent", digits = 3)
```

```{r metric-distribution-stack, fig.width=7.5, fig.height=3.2}
agent_colors <- c("User" = "#377EB8", "LLM" = "#FF7F00")

df_metric_distributions <- df_res_long %>%
  select(agent, surprise, transience, resonance) %>%
  pivot_longer(
    cols = c(surprise, transience, resonance),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c("surprise", "transience", "resonance"),
      labels = c("Novelty", "Transience", "Resonance")
    )
  )

metric_summary <- df_metric_distributions %>%
  group_by(metric, agent) %>%
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    .groups = "drop"
  )

p_metric_distributions <- ggplot(
  df_metric_distributions,
  aes(x = agent, y = value, fill = agent)
) +
  geom_violin(alpha = 0.4, colour = NA, trim = FALSE, na.rm = TRUE) +
  geom_boxplot(width = 0.18, outlier.shape = NA, alpha = 0.9, na.rm = TRUE) +
  facet_wrap(~ metric, nrow = 1, scales = "free_y") +
  scale_fill_manual(values = agent_colors, name = NULL) +
  scale_color_manual(values = agent_colors, name = NULL) +
  labs(
    title = "Novelty Metrics by Agent",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 10),
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.margin = margin(4, 4, 4, 4)
  )

print(p_metric_distributions)

ggsave(
  file.path(GRAPHS_DIR, "Novelty_Transience_Resonance_Distributions_Agent.png"),
  p_metric_distributions,
  width = 7.5,
  height = 3.2,
  dpi = 300
)
```


# 2. Novelty and Resonance Analysis

## 2.1 Resonance vs Novelty Plot

```{r resonance-plot, fig.width=12, fig.height=6}
p_resonance <- ggplot(df_res_long, aes(x = surprise, y = resonance, color = agent)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE, size = 1.2) +
  scale_color_manual(values = c("User" = "#377EB8", "LLM" = "#FF7F00")) +
  geom_vline(data = desc_stats,
             aes(xintercept = mean_surprise, color = agent),
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  labs(
    title = "Resonance vs Novelty (Field Data)",
    x = "Novelty",
    y = "Resonance (Novelty - Transience)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top") + xlim(-5, 1)

print(p_resonance)

ggsave(file.path(GRAPHS_DIR, "novelty_resonance_vs_novelty.png"),
       p_resonance, width = 12, height = 6, dpi = 300)
```

## 2.2 Transience vs Novelty Plot

```{r transience-plot, fig.width=12, fig.height=6}
p_transience <- ggplot(df_res_long, aes(x = surprise, y = transience, color = agent)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE, size = 1.2) +
  scale_color_manual(values = c("User" = "#377EB8", "LLM" = "#FF7F00")) +
  geom_vline(data = desc_stats,
             aes(xintercept = mean_surprise, color = agent),
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  labs(
    title = "Transience vs Novelty (Field Data)",
    x = "Novelty (Surprisal)",
    y = "Transience"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")

print(p_transience)

ggsave(file.path(GRAPHS_DIR, "novelty_transience_vs_novelty.png"),
       p_transience, width = 12, height = 6, dpi = 300)
```

# 3. Statistical Models

## 3.1 Resonance Model

```{r resonance-model}
model_resonance <- lmer(resonance ~ surprise * agent + (1 | conversation_id), data = df_res_long)
cat("=== Resonance Model ===\n")
print(summary(model_resonance)$coefficients)

cat("\n=== Slope Comparison (Resonance) ===\n")
print(emtrends(model_resonance, specs = "agent", var = "surprise"))

model_resonance_len <- lmer(resonance ~ surprise * agent + nwords + (1 | conversation_id), data = df_res_long)
cat("\n=== Resonance Model (controlling for turn length) ===\n")
print(summary(model_resonance_len)$coefficients)

cat("\n=== Slope Comparison (Resonance, length-controlled) ===\n")
print(emtrends(model_resonance_len, specs = "agent", var = "surprise"))
```

## 3.2 Transience Model

```{r transience-model}
model_transience <- lmer(transience ~ surprise * agent + (1 | conversation_id), data = df_res_long)

cat("=== Transience Model ===\n")
print(summary(model_transience)$coefficients)

cat("\n=== Slope Comparison (Transience) ===\n")
print(emtrends(model_transience, specs = "agent", var = "surprise"))

model_transience_len <- lmer(transience ~ surprise * agent + nwords + (1 | conversation_id), data = df_res_long)

cat("\n=== Transience Model (controlling for turn length) ===\n")
print(summary(model_transience_len)$coefficients)

cat("\n=== Slope Comparison (Transience, length-controlled) ===\n")
print(emtrends(model_transience_len, specs = "agent", var = "surprise"))
```

# 4. Supplementary: Resonance vs Novelty by LLM Type

```{r llm-resonance-novelty, fig.width=12, fig.height=6}
desc_stats_llm <- df_res_long %>%
  filter(!is.na(llm_type)) %>%
  group_by(agent, llm_type) %>%
  summarise(mean_surprise = mean(surprise, na.rm = TRUE), .groups = "drop")

p_llm_user <- df_res_long %>%
  filter(!is.na(llm_type), agent == "User") %>%
  ggplot(aes(x = surprise, y = resonance, color = factor(llm_type))) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  geom_vline(data = desc_stats_llm %>% filter(agent == "User"),
             aes(xintercept = mean_surprise, color = factor(llm_type)),
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  labs(
    title  = "User: Resonance vs Novelty by LLM Type",
    x      = "Novelty",
    y      = "Resonance (Novelty − Transience)",
    color  = "LLM Type"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top",
        plot.title = element_text(hjust = 0.5, face = "bold"))

p_llm_ai <- df_res_long %>%
  filter(!is.na(llm_type), agent == "LLM") %>%
  ggplot(aes(x = surprise, y = resonance, color = factor(llm_type))) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  geom_vline(data = desc_stats_llm %>% filter(agent == "LLM"),
             aes(xintercept = mean_surprise, color = factor(llm_type)),
             linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
  labs(
    title  = "LLM: Resonance vs Novelty by LLM Type",
    x      = "Novelty",
    y      = "Resonance (Novelty − Transience)",
    color  = "LLM Type"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top",
        plot.title = element_text(hjust = 0.5, face = "bold"))

print(p_llm_user)
print(p_llm_ai)

ggsave(file.path(GRAPHS_DIR, "novelty_resonance_llm_type_user.png"),
       p_llm_user, width = 12, height = 6, dpi = 300)
ggsave(file.path(GRAPHS_DIR, "novelty_resonance_llm_type_ai.png"),
       p_llm_ai, width = 12, height = 6, dpi = 300)
```

# 5. Session Information

```{r session-info}
sessionInfo()
```
