---
title: "Robustness Checks"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
source(here::here("analysis/comparison/_comparison_utils.R"))
pacman::p_load(vegan)
```

The robustness checks in the paper's *Robustness and Permutation* section.
The scorer-familiarity check (AA stratified by model family) is in
`transience_uptake_and_scorer_checks.Rmd`.

# Shuffled context (Human-AI)

Each HA turn is scored with no context ($s_{\text{base}}$), with its own
story's context ($s_{\text{true}}$) and with the context of another, randomly
assigned HA story ($s_{\text{shuffled}}$), using `google/gemma-4-12B`
(`scripts/shuffled_context_control.py`). Gains are reductions in surprisal
(bits/token): $G_{\text{true}} = s_{\text{base}} - s_{\text{true}}$,
$G_{\text{shuffled}} = s_{\text{base}} - s_{\text{shuffled}}$ and
$G_{\text{content}} = G_{\text{true}} - G_{\text{shuffled}}$. The scorer
differs from the main analysis (`gemma-4-31B`), so these gains are compared
with each other, not with the main novelty scores.

```{r shuffled}
shuffled <- read_csv(here("data/human-ai/processed/shuffled_control_results.csv"), show_col_types = FALSE)

gains <- shuffled %>%
  group_by(speaker) %>%
  summarise(n_turns = n(), across(c(gain_true, gain_shuffled, gain_content), mean), .groups = "drop")

gains %>%
  knitr::kable(digits = 3, caption = "Mean surprisal gain by speaker (bits/token, all scored turns)")

gap <- function(col) gains[[col]][gains$speaker == "ai"] - gains[[col]][gains$speaker == "human"]
tibble(
  gain = c("true", "shuffled", "content"),
  ai_minus_human = c(gap("gain_true"), gap("gain_shuffled"), gap("gain_content"))
) %>%
  mutate(share_of_true_gap = ai_minus_human / gap("gain_true")) %>%
  knitr::kable(digits = 3, caption = "AI - human gap in gain")

per_story <- shuffled %>%
  group_by(conversation_id, speaker) %>%
  summarise(gain_content = mean(gain_content), .groups = "drop") %>%
  pivot_wider(names_from = speaker, values_from = gain_content) %>%
  drop_na(ai, human)

test <- t.test(per_story$ai, per_story$human, paired = TRUE)
cat(sprintf(
  "Per-story paired t-test, G_content AI vs human: mean diff = %.3f, t(%d) = %.2f, p = %.2g\n",
  unname(test$estimate), as.integer(test$parameter), unname(test$statistic), test$p.value
))
```

# Simpler lexical and length baselines (Human-AI)

Per turn, on content words (stopwords removed):

- **new-word rate**: share of the turn's words not used earlier in the story
- **lexical uptake**: overlap with the immediately preceding turn, as Jaccard
  similarity and as echo-of-prior (share of the previous turn's words reused).
  Echo-of-prior normalizes by the previous turn, so it does not grow with the
  length of the current turn.
- **turn length**: number of content words

Each measure is averaged over human and over AI turns within a story and
compared with a paired t-test (n = stories); $d_z$ is the standardized paired
difference, human minus AI. Surprisal novelty (the paper's metric, turns with
$1 < $ `analysis_turn` $\le 9$) is included for reference.

```{r lexical}
stopwords <- c(
  "the", "a", "an", "and", "or", "but", "if", "of", "to", "in", "on", "at", "for", "with", "as", "by", "from",
  "is", "are", "was", "were", "be", "been", "being", "it", "its", "this", "that", "these", "those", "i", "you",
  "he", "she", "they", "we", "him", "her", "them", "his", "their", "our", "my", "me", "your", "s", "t", "not",
  "no", "so", "then", "than", "too", "very", "can", "will", "would", "could", "should", "had", "has", "have",
  "do", "does", "did", "up", "out", "about", "into", "over", "after", "just", "who", "what", "when", "where",
  "which", "there", "here", "all", "any", "some", "more", "most", "one", "like", "also", "only", "d", "re",
  "ll", "m", "ve"
)
tokens <- function(x) {
  w <- unlist(strsplit(tolower(gsub("[^a-z0-9' ]", " ", tolower(x))), " +"))
  w <- w[w != ""]
  w[!(w %in% stopwords)]
}
jaccard <- function(a, b) {
  u <- length(union(a, b))
  if (u == 0) NA_real_ else length(intersect(a, b)) / u
}
echo_of_prior <- function(cur, prev) if (length(prev) == 0) NA_real_ else length(intersect(cur, prev)) / length(prev)

# Base read.csv keeps empty turns as "" rather than NA, as in the original analysis.
ha <- read.csv(here("data/human-ai/processed/novelty_scores.csv"), stringsAsFactors = FALSE)
ha$conversation_id <- clean_conversation_id(ha$conversation_id)
complete <- tolower(as.character(ha$complete_exchange))
ha <- ha[(is.na(ha$complete_exchange) | complete == "true") & !(ha$conversation_id %in% EXCLUDED_CONVERSATIONS), ]
ha <- ha[order(ha$conversation_id, ha$turn), ]

lexical <- do.call(rbind, lapply(unique(ha$conversation_id), function(cid) {
  g <- ha[ha$conversation_id == cid, ]
  stream <- list()
  who <- character(0)
  for (i in seq_len(nrow(g))) {
    stream[[length(stream) + 1]] <- tokens(g$author_1[i]); who <- c(who, "H")
    stream[[length(stream) + 1]] <- tokens(g$author_2[i]); who <- c(who, "A")
  }
  n <- length(stream)
  new_rate <- jac <- echo <- len <- rep(NA_real_, n)
  seen <- character(0)
  for (i in seq_len(n)) {
    cur <- stream[[i]]
    len[i] <- length(cur)
    if (i > 1 && length(cur) > 0) {
      new_rate[i] <- mean(!(cur %in% seen))
      jac[i] <- jaccard(cur, stream[[i - 1]])
      echo[i] <- echo_of_prior(cur, stream[[i - 1]])
    }
    seen <- union(seen, cur)
  }
  data.frame(
    conversation_id = cid,
    h_newrate = mean(new_rate[who == "H"], na.rm = TRUE), a_newrate = mean(new_rate[who == "A"], na.rm = TRUE),
    h_jac = mean(jac[who == "H"], na.rm = TRUE), a_jac = mean(jac[who == "A"], na.rm = TRUE),
    h_echo = mean(echo[who == "H"], na.rm = TRUE), a_echo = mean(echo[who == "A"], na.rm = TRUE),
    h_len = mean(len[who == "H"]), a_len = mean(len[who == "A"])
  )
}))

window <- ha[!is.na(ha$analysis_turn) & ha$analysis_turn > 1 & ha$analysis_turn <= 9, ]
novelty <- aggregate(cbind(author_1_surprise, author_2_surprise) ~ conversation_id, window, mean)
names(novelty) <- c("conversation_id", "h_nov", "a_nov")
lexical <- merge(lexical, novelty, by = "conversation_id")

paired <- function(name, h, a) {
  t <- t.test(h, a, paired = TRUE)
  data.frame(measure = name, human = mean(h), ai = mean(a), dz = mean(h - a) / sd(h - a), p = t$p.value,
             higher = ifelse(mean(h) > mean(a), "human", "AI"))
}
rbind(
  paired("surprisal novelty (paper)", lexical$h_nov, lexical$a_nov),
  paired("new-word rate", lexical$h_newrate, lexical$a_newrate),
  paired("lexical uptake: Jaccard", lexical$h_jac, lexical$a_jac),
  paired("lexical uptake: echo-of-prior", lexical$h_echo, lexical$a_echo),
  paired("turn length (content words)", lexical$h_len, lexical$a_len)
) %>%
  knitr::kable(digits = c(0, 3, 3, 2, 4, 0), caption = sprintf("Human vs AI per story (n = %d)", nrow(lexical)))

novelty_asym <- lexical$h_nov - lexical$a_nov
convergence <- function(name, x) {
  s <- suppressWarnings(cor.test(x, novelty_asym, method = "spearman"))
  data.frame(asymmetry = name, spearman_rho = unname(s$estimate), p = s$p.value)
}
rbind(
  convergence("new-word rate (human - AI)", lexical$h_newrate - lexical$a_newrate),
  convergence("turn length (AI - human)", lexical$a_len - lexical$h_len)
) %>%
  knitr::kable(digits = 3, caption = "Correlation of simple asymmetries with the surprisal-novelty asymmetry across stories")
```

# Surface metrics PERMANOVA (HA, HH, AA)

For five textdescriptives metrics, each story contributes the absolute
difference between its two slot means, $A_m(i)$, over complete exchanges with
$1 <$ `analysis_turn` $\le 9$. The five-dimensional $A$ vectors are compared
across conditions with PERMANOVA (Euclidean, 9,999 permutations), and per
metric with the planned HA vs (HH + AA)/2 contrast.

```{r surface}
surface_metrics <- c(
  word_count = "n_tokens",
  sentence_length = "sentence_length_mean",
  lexical_diversity = "proportion_unique_tokens",
  readability = "flesch_reading_ease",
  coherence = "first_order_coherence"
)

surface <- load_surface_metrics(PUBLISHED_CONDITIONS) %>%
  filter_complete_exchange_window() %>%
  filter(turn_index > 1) %>%
  prepare_condition_factor(reference = "human-ai") %>%
  filter(!conversation_id %in% c("conv_ed575a06c11d42358e3eeb7826d2f959", "conv_a63a08273d0a4704a7638e4cd6850225")) %>%
  filter(speaker_slot %in% c("author_1", "author_2")) %>%
  mutate(slot = factor(speaker_slot, levels = c("author_1", "author_2"), labels = c("Slot 1", "Slot 2")))

surface_A <- purrr::imap_dfr(surface_metrics, function(column, metric) {
  surface %>%
    select(condition, conversation_id, slot, value = all_of(column)) %>%
    group_by(condition, conversation_id, slot) %>%
    summarise(side_mean = mean(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = slot, values_from = side_mean) %>%
    drop_na(`Slot 1`, `Slot 2`) %>%
    transmute(metric, condition, conversation_id, A = abs(`Slot 1` - `Slot 2`))
})

surface_A %>%
  group_by(metric, condition) %>%
  summarise(n = n(), mean_A = mean(A), median_A = median(A), .groups = "drop") %>%
  knitr::kable(digits = 3, caption = "Per-story slot asymmetry A_m by condition")

surface_A %>%
  group_by(metric) %>%
  group_modify(~ planned_contrast_HA_vs_same(.x, "A")) %>%
  ungroup() %>%
  mutate(p_BH = p.adjust(p, method = "BH")) %>%
  select(metric, estimate, ci_lower, ci_upper, p, p_BH) %>%
  knitr::kable(digits = 3, caption = "Planned contrast HA vs (HH + AA)/2 per metric, BH-adjusted across metrics")

A_matrix <- surface_A %>%
  pivot_wider(id_cols = c(conversation_id, condition), names_from = metric, values_from = A) %>%
  drop_na()
cat(sprintf("Stories with all %d metrics: %d\n", length(surface_metrics), nrow(A_matrix)))

set.seed(42)
adonis2(
  as.matrix(select(A_matrix, -conversation_id, -condition)) ~ condition,
  data = A_matrix, method = "euclidean", permutations = 9999
)
```
