---
title: "Semantic Distance Across Conditions"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

source(here::here("analysis/comparison/_comparison_utils.R"))
CONDITIONS <- ALL_CONDITIONS  # or PUBLISHED_CONDITIONS for the three-condition paper set

pacman::p_load(lme4, lmerTest, patchwork, emmeans)
GRAPHS_DIR <- file.path(PROJECT_ROOT, "analysis", "figures")

theme_set(theme_comparison())
```

# Overview

`semantic_exploration_binned.parquet` holds lag-based distances: `k` is the lag, `distance` the mean cosine distance at that lag, and `agent` is `"interleaved"` (full exchange sequence) or `author_1` / `author_2`.

This notebook uses the turn embeddings to reconstruct the alternating sequence and analyse:

1. **Adjacent turn-to-turn semantic distance** across conditions as the primary analysis
2. **Story-level mean adjacent distance** as the supporting aggregate
3. **Turn-level adjacent semantic-distance networks** as descriptive 2D maps
4. **Lag-based trajectories** as supplemental descriptive context

- The adjacent sequence includes both within-exchange steps (e.g., slot 1$_t$ to slot 2$_t$) and between-exchange steps (e.g., slot 2$_t$ to slot 1$_{t+1}$).
- For Human-AI, the sequence is kept in the standard human-to-AI row order. For same-type dyads, `starter_side` determines which slot speaks first within each exchange.
- **Human / AI** in Human-AI and **Starter-side / Non-starter-side** in same-type conditions are only used as descriptive views.

**Primary cross-condition object.** Each adjacent pair in the reconstructed alternating sequence contributes one semantic distance. The main model compares these adjacent distances by condition while nesting observations within story:
$$d_{ij} = \beta_0 + \beta_1 \mathrm{condition}_{ij} + u_{\mathrm{story}(j)} + \epsilon_{ij}.$$

The 2D network layouts are precomputed by `scripts/07_compute_semantic_network_layouts.py`, which samples the same number of stories and edges per condition.

# Load Data

```{r load-data}
excluded_conversations <- c(
  "conv_ed575a06c11d42358e3eeb7826d2f959",
  "conv_a63a08273d0a4704a7638e4cd6850225"
)

df_combined <- load_exploration_data(CONDITIONS) %>%
  filter(k > 1) %>%
  prepare_condition_factor(reference = "human-ai") %>%
  filter(!conversation_id %in% excluded_conversations)

df_embeddings <- load_conditions(
  CONDITIONS,
  "story_embeddings_interaction_level.parquet",
  prepare = function(df) df %>% filter_complete_exchange_window() %>% drop_provider_error_stories()
) %>%
  filter(!conversation_id %in% excluded_conversations) %>%
  prepare_condition_factor(reference = "human-ai")

df_combined %>%
  group_by(condition, agent) %>%
  summarise(
    n_stories = n_distinct(conversation_id),
    n_rows = n(),
    min_k = min(k, na.rm = TRUE),
    max_k = max(k, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Loaded semantic exploration rows by condition and agent view")
```

# Preprocessing

```{r preprocess}
df_interleaved <- df_combined %>%
  filter(agent == "interleaved") %>%
  drop_na(distance, k)

df_speaker <- df_combined %>%
  filter(agent != "interleaved") %>%
  filter(speaker_slot %in% c("author_1", "author_2")) %>%
  mutate(
    slot = factor(speaker_slot, levels = c("author_1", "author_2"), labels = c("Slot 1", "Slot 2"))
  ) %>%
  drop_na(distance, k)

cat("Interleaved rows:", nrow(df_interleaved), "\n")
cat("Speaker rows:", nrow(df_speaker), "\n")
cat("Embedding rows:", nrow(df_embeddings), "\n")
```

# Primary: Adjacent Turn-to-Turn Semantic Distance

```{r adjacent-distance-functions}
parse_embedding_r <- function(x) {
  if (is.null(x) || length(x) == 0) return(NULL)
  if (is.character(x)) {
    x <- gsub("^\\[|\\]$", "", x)
    out <- suppressWarnings(as.numeric(strsplit(x, ",")[[1]]))
  } else {
    out <- suppressWarnings(as.numeric(unlist(x, use.names = FALSE)))
  }
  if (length(out) == 0 || any(!is.finite(out))) return(NULL)
  out
}

cosine_distance <- function(a, b) {
  if (is.null(a) || is.null(b) || length(a) != length(b)) return(NA_real_)
  denom <- sqrt(sum(a * a)) * sqrt(sum(b * b))
  if (!is.finite(denom) || denom == 0) return(NA_real_)
  sim <- sum(a * b) / denom
  1 - pmax(pmin(sim, 1), -1)
}

story_slot_order <- function(story_df) {
  condition_value <- as.character(story_df$condition[1])
  starter_values <- na.omit(as.character(story_df$starter_side))
  starter_side <- if (length(starter_values) > 0) starter_values[1] else NA_character_

  if (condition_value == "Human-AI") {
    c("author_1", "author_2")
  } else if (!is.na(starter_side) && starter_side == "author_2") {
    c("author_2", "author_1")
  } else {
    c("author_1", "author_2")
  }
}

build_story_sequence <- function(story_df) {
  story_df <- story_df %>%
    arrange(turn_index)

  slot_order <- story_slot_order(story_df)

  map_dfr(seq_len(nrow(story_df)), function(i) {
    row_i <- story_df[i, ]
    first_slot <- slot_order[1]
    second_slot <- slot_order[2]

    tibble(
      sequence_index = c(2 * i - 1, 2 * i),
      turn_index = c(row_i$turn_index, row_i$turn_index),
      slot = c(first_slot, second_slot),
      embedding = list(
        parse_embedding_r(row_i[[paste0(first_slot, "_embedding")]][[1]]),
        parse_embedding_r(row_i[[paste0(second_slot, "_embedding")]][[1]])
      )
    )
  }) %>%
    filter(map_lgl(embedding, ~ !is.null(.x)))
}

build_story_adjacent_distances <- function(story_df) {
  sequence_df <- build_story_sequence(story_df)

  if (nrow(sequence_df) < 2) {
    return(tibble())
  }

  n_seq <- nrow(sequence_df)

  tibble(
    pair_index = seq_len(n_seq - 1),
    from_turn = sequence_df$turn_index[-n_seq],
    to_turn = sequence_df$turn_index[-1],
    from_slot = sequence_df$slot[-n_seq],
    to_slot = sequence_df$slot[-1],
    pair_type = if_else(from_turn == to_turn, "within_exchange", "between_exchange"),
    adjacent_distance = map2_dbl(
      sequence_df$embedding[-n_seq],
      sequence_df$embedding[-1],
      cosine_distance
    )
  ) %>%
    filter(is.finite(adjacent_distance))
}

df_adjacent <- df_embeddings %>%
  group_by(condition, conversation_id) %>%
  group_modify(~ build_story_adjacent_distances(.x), .keep = TRUE) %>%
  ungroup() %>%
  mutate(
    pair_type = factor(pair_type, levels = c("within_exchange", "between_exchange"))
  )

df_adjacent %>%
  group_by(condition) %>%
  summarise(
    n_adjacent_pairs = n(),
    n_stories = n_distinct(conversation_id),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Adjacent semantic-distance observations by condition")

df_adjacent %>%
  group_by(condition) %>%
  summarise(
    n = n(),
    n_stories = n_distinct(conversation_id),
    mean_distance = mean(adjacent_distance),
    sd_distance = sd(adjacent_distance),
    median_distance = median(adjacent_distance),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Adjacent turn-to-turn semantic distance by condition"
  )
```

```{r adjacent-distance-model}
if (n_distinct(df_adjacent$condition) >= 2 && nrow(df_adjacent) >= 10) {
  model_adjacent <- lmer(
    adjacent_distance ~ condition + (1 | conversation_id),
    data = df_adjacent
  )

  cat("Main mixed model: adjacent_distance ~ condition + (1 | conversation_id)\n")
  print(summary(model_adjacent))

  emm_adjacent <- emmeans(model_adjacent, specs = "condition")

  cat("\nEstimated marginal means by condition:\n")
  print(emm_adjacent)

  cat("\nPairwise condition contrasts (BH-adjusted):\n")
  print(pairs(emm_adjacent, adjust = "BH"))

  cat("\nPlanned contrast: Human-AI vs same-type average:\n")
  print(contrast(
    emm_adjacent,
    method = list("Human-AI - mean(Human-Human, AI-AI)" = condition_weights(emm_adjacent, HA_VS_SAME_TYPE))
  ))
}
```

```{r adjacent-distance-plot, fig.width=9, fig.height=4.5}
semantic_density_boxplot <- function(df, value_col, title, subtitle, x_label) {
  plot_data <- df %>%
    select(condition, value = all_of(value_col)) %>%
    drop_na(value)

  density_panel <- ggplot(plot_data, aes(x = value, fill = condition, color = condition)) +
    geom_density(alpha = 0.28, linewidth = 0.9, na.rm = TRUE) +
    scale_fill_manual(values = condition_colors, name = NULL) +
    scale_color_manual(values = condition_colors, name = NULL) +
    labs(
      title = title,
      subtitle = subtitle,
      x = NULL,
      y = "Density"
    ) +
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 10),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      plot.margin = margin(4, 8, 0, 8)
    )

  box_panel <- ggplot(plot_data, aes(x = value, y = condition, fill = condition)) +
    geom_boxplot(width = 0.45, outlier.alpha = 0.2, alpha = 0.9, na.rm = TRUE) +
    scale_fill_manual(values = condition_colors, name = NULL) +
    labs(x = x_label, y = NULL) +
    theme(
      legend.position = "none",
      axis.text.y = element_text(size = 10),
      plot.margin = margin(0, 8, 4, 8)
    )

  density_panel / box_panel + plot_layout(heights = c(2.6, 1))
}

turn_level_semanticplots <- semantic_density_boxplot(
  df_adjacent,
  "adjacent_distance",
  "Adjacent turn-to-turn semantic distance by condition",
  "Each density overlays one condition; boxplots summarize the same values below",
  "Adjacent semantic distance"
)

print(turn_level_semanticplots)

ggsave(
  file.path(GRAPHS_DIR, "semantic_distance_interactionlevel.png"),
  turn_level_semanticplots,
  width = 9,
  height = 4.5,
  dpi = 300
)

```

# Supporting: Story-Level Mean Adjacent Distance

```{r story-mean-adjacent}
df_adjacent_story <- df_adjacent %>%
  group_by(condition, conversation_id) %>%
  summarise(
    n_pairs = n(),
    mean_adjacent_distance = mean(adjacent_distance, na.rm = TRUE),
    median_adjacent_distance = median(adjacent_distance, na.rm = TRUE),
    .groups = "drop"
  )

df_adjacent_story %>%
  group_by(condition) %>%
  summarise(
    n = n(),
    mean_story_distance = mean(mean_adjacent_distance),
    sd_story_distance = sd(mean_adjacent_distance),
    median_story_distance = median(mean_adjacent_distance),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Story-level mean adjacent semantic distance by condition"
  )
```

```{r story-mean-adjacent-tests}
if (n_distinct(df_adjacent_story$condition) >= 2 && nrow(df_adjacent_story) >= 10) {
  cat("One-way ANOVA: mean_adjacent_distance ~ condition\n")
  print(summary(aov(mean_adjacent_distance ~ condition, data = df_adjacent_story)))

  cat("\nKruskal-Wallis (robust): mean_adjacent_distance ~ condition\n")
  print(kruskal.test(mean_adjacent_distance ~ condition, data = df_adjacent_story))

  cat("\nPairwise Wilcoxon (BH-adjusted):\n")
  print(pairwise.wilcox.test(
    df_adjacent_story$mean_adjacent_distance,
    df_adjacent_story$condition,
    p.adjust.method = "BH"
  ))
}
```

```{r story-mean-adjacent-plot, fig.width=9, fig.height=4.5}
semantic_distance_storylevel <- semantic_density_boxplot(
  df_adjacent_story,
  "mean_adjacent_distance",
  "Story-level mean adjacent semantic distance by condition",
  "One mean adjacent distance per story",
  "Mean adjacent semantic distance"
)

print(semantic_distance_storylevel)

ggsave(
  file.path(GRAPHS_DIR, "semantic_distance_storylevel.png"),
  semantic_distance_storylevel,
  width = 9,
  height = 4.5,
  dpi = 300
)
```

# Bootstrap CIs and Effect Sizes on Story Means

```{r boot-effects-adjacent-story}
bootstrap_condition_means(df_adjacent_story, "mean_adjacent_distance", R = 10000) %>%
  knitr::kable(
    digits = 3,
    caption = "Bootstrap 95% CI on per-condition mean of story-level adjacent distance (R = 10,000)"
  )

effect_sizes_pairwise(df_adjacent_story, "mean_adjacent_distance") %>%
  knitr::kable(
    digits = 3,
    caption = "Pairwise Cliff's delta on story-level mean adjacent distance"
  )
```

# Planned Contrast: HA vs (HH + AA)/2

```{r planned-contrast-adjacent-story}
planned_contrast_HA_vs_same(df_adjacent_story, "mean_adjacent_distance") %>%
  knitr::kable(
    digits = 3,
    caption = "Planned contrast: story-level mean adjacent distance in HA vs the average of HH and AA"
  )
```

# Supplemental: Turn-Level Adjacent Semantic-Distance Network

```{r turn-adjacent-network-data}
network_dir <- here::here("analysis", "comparison", "semantic_network_layouts")
network_paths <- list(
  turn_nodes = file.path(network_dir, "turn_nodes.csv"),
  turn_edges = file.path(network_dir, "turn_edges.csv"),
  metadata = file.path(network_dir, "metadata.json")
)

network_available <- all(file.exists(unlist(network_paths)))

if (network_available) {
  df_turn_network <- read_csv(network_paths$turn_nodes, show_col_types = FALSE) %>%
    mutate(condition = factor(condition, levels = names(condition_colors)))
  turn_edges <- read_csv(network_paths$turn_edges, show_col_types = FALSE) %>%
    mutate(
      condition = factor(condition, levels = names(condition_colors)),
      adjacent_distance = as.numeric(adjacent_distance),
      closeness = 1 - scales::rescale(
        adjacent_distance,
        to = c(0, 1),
        from = quantile(adjacent_distance, c(0.05, 0.95), na.rm = TRUE)
      ),
      closeness = pmax(pmin(closeness, 1), 0),
      pair_type = factor(pair_type, levels = c("within_exchange", "between_exchange"))
    )
  network_metadata <- jsonlite::fromJSON(network_paths$metadata)
  close_distance_threshold <- quantile(turn_edges$adjacent_distance, 0.40, na.rm = TRUE)
  close_edges <- turn_edges %>%
    filter(adjacent_distance <= close_distance_threshold)

  turn_edge_labels <- turn_edges %>%
    group_by(condition) %>%
    summarise(
      mean_distance = mean(adjacent_distance, na.rm = TRUE),
      x = quantile(c(x, xend), 0.03, na.rm = TRUE),
      y = quantile(c(y, yend), 0.97, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(
      close_edges %>%
        count(condition, name = "n_close"),
      by = "condition"
    ) %>%
    mutate(
      pct_close = 100 * n_close / as.numeric(table(turn_edges$condition)[as.character(condition)]),
      label = sprintf("Mean d = %.3f\nClose edges: %.0f%%", mean_distance, pct_close)
    )
} else {
  df_turn_network <- tibble()
  turn_edges <- tibble()
  close_edges <- tibble()
  turn_edge_labels <- tibble()
  close_distance_threshold <- NA_real_
  network_metadata <- list(turn_layout_method = "not computed")
  cat(
    "Precomputed semantic network files not found. Run:\n",
    "python scripts/07_compute_semantic_network_layouts.py\n"
  )
}
```

```{r turn-adjacent-network-plot, fig.width=13, fig.height=6}
if (network_available) {
  network_panel <- ggplot() +
    geom_point(
      data = df_turn_network,
      aes(x = x, y = y),
      color = "grey65",
      alpha = 0.18,
      size = 0.45
    ) +
    geom_segment(
      data = close_edges,
      aes(
        x = x,
        y = y,
        xend = xend,
        yend = yend,
        linewidth = closeness,
        group = conversation_id
      ),
      color = "grey10",
      alpha = 0.58,
      lineend = "round"
    ) +
    geom_label(
      data = turn_edge_labels,
      aes(x = x, y = y, label = label),
      hjust = 0,
      vjust = 1,
      label.size = 0,
      fill = "white",
      alpha = 0.85,
      size = 3.2
    ) +
    facet_wrap(~condition, nrow = 1) +
    scale_linewidth_continuous(range = c(0.15, 1.55), guide = "none") +
    labs(
      title = "Adjacent semantic continuity networks",
      subtitle = paste0(
        "Only semantically close links (global closest 40%; d <= ",
        sprintf("%.3f", close_distance_threshold),
        "). Missing links indicate larger semantic jumps"
      ),
      x = NULL,
      y = NULL
    ) +
    theme(
      legend.position = "none",
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      panel.background = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA)
    )

  distance_panel <- ggplot(turn_edges, aes(x = adjacent_distance, fill = condition)) +
    geom_density(alpha = 0.75, color = NA, adjust = 1.1) +
    geom_vline(
      xintercept = close_distance_threshold,
      linetype = "dashed",
      color = "grey25",
      linewidth = 0.35
    ) +
    facet_wrap(~condition, nrow = 1) +
    scale_fill_manual(values = condition_colors) +
    labs(
      x = "Adjacent semantic distance",
      y = "Density"
    ) +
    theme(
      legend.position = "none",
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA),
      strip.text = element_blank()
    )

  semantic_network <- network_panel / distance_panel +
    plot_layout(heights = c(3.2, 1))
} else {
  semantic_network <- ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = "Run python scripts/07_compute_semantic_network_layouts.py to generate this plot."
    ) +
    theme_void()
}

ggsave(
  file.path(GRAPHS_DIR, "semantic_network.png"),
  semantic_network,
  width = 12,
  height = 8,
  dpi = 300
)

```


# Supplemental: General Condition-Level Comparison

```{r interleaved-descriptives}
df_interleaved %>%
  group_by(condition, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    sd_distance = sd(distance, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Interleaved semantic distance by lag and condition"
  )
```

```{r interleaved-plot, fig.width=12, fig.height=5}
interleaved_summary <- df_interleaved %>%
  group_by(condition, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(interleaved_summary, aes(x = k, y = mean_distance, color = condition)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_ribbon(
    aes(ymin = mean_distance - 1.96 * se_distance,
        ymax = mean_distance + 1.96 * se_distance,
        fill = condition),
    alpha = 0.18,
    color = NA
  ) +
  scale_color_manual(values = condition_colors) +
  scale_fill_manual(values = condition_colors) +
  labs(
    title = "Interleaved Semantic Exploration by Lag",
    subtitle = "Mean distance with 95% confidence intervals",
    x = "Lag Bin (k)",
    y = "Semantic Distance"
  ) +
  theme(legend.position = "bottom")
```

```{r interleaved-model}
model_interleaved <- lmer(
  distance ~ k * condition + (1 | conversation_id),
  data = df_interleaved
)

cat("Interleaved model summary:\n")
print(summary(model_interleaved))

emm_interleaved <- emmeans(model_interleaved, specs = "condition")

cat("\nCondition contrasts (interleaved view):\n")
print(pairs(emm_interleaved, adjust = "bonferroni"))
```

# Supplemental: Role-Level View

```{r speaker-descriptives}
df_speaker %>%
  group_by(condition, condition_role, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Speaker-level semantic distance by condition, role, and lag"
  )
```

```{r role-plot, fig.width=12, fig.height=6}
role_summary <- df_speaker %>%
  filter(!is.na(condition_role)) %>%
  group_by(condition, condition_role, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    se_distance = sd(distance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(role_summary, aes(x = k, y = mean_distance, color = condition_role)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  geom_ribbon(
    aes(ymin = mean_distance - 1.96 * se_distance,
        ymax = mean_distance + 1.96 * se_distance,
        fill = condition_role),
    alpha = 0.15,
    color = NA
  ) +
  facet_wrap(~condition) +
  labs(
    title = "Speaker-Level Semantic Exploration by Role",
    subtitle = "Within-condition role view only: Human / AI in Human-AI; Starter-side / Non-starter-side in same-type conditions",
    x = "Lag Bin (k)",
    y = "Semantic Distance",
    color = "Role",
    fill = "Role"
  ) +
  theme(legend.position = "bottom")
```

# Supplemental: Interleaved vs Speaker View

```{r view-comparison, fig.width=12, fig.height=5}
view_summary <- bind_rows(
  df_interleaved %>%
    mutate(view = "Interleaved sequence") %>%
    select(condition, conversation_id, k, distance, view),
  df_speaker %>%
    mutate(view = "Speaker-specific") %>%
    select(condition, conversation_id, k, distance, view)
) %>%
  group_by(condition, view, k) %>%
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(view_summary, aes(x = k, y = mean_distance, color = condition)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  facet_wrap(~view) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Interleaved vs Speaker-Specific Semantic Distance",
    x = "Lag Bin (k)",
    y = "Mean Semantic Distance"
  ) +
  theme(legend.position = "bottom")
```

# Summary

Primary cross-condition result: adjacent turn-to-turn semantic distance from the reconstructed alternating dyadic sequence, tested with a mixed model that nests adjacent observations within story. This includes both within-exchange distances and the cross-exchange step from the second turn of one exchange to the first turn of the next exchange.

Robustness layers added:

- story-level mean adjacent distance as an aggregate supporting analysis
- bootstrap 95% CIs on per-condition story means
- pairwise Cliff's $\delta$ effect sizes on story means
- planned single-df contrast HA vs (HH+AA)/2 on story means

Supplemental views retained:

- turn-level adjacent semantic-distance network, with edge color/width encoding original-space adjacent distance
- lag-based speaker-level slot trajectories
- lag-based interleaved sequence model and condition descriptives
- within-condition role curves (descriptive only)
- interleaved vs speaker-specific side-by-side view
