---
title: "Transience uptake (2x2) and scorer-familiarity checks"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

source(here::here("analysis/comparison/_comparison_utils.R"))
CONDITIONS <- ALL_CONDITIONS  # or PUBLISHED_CONDITIONS for the three-condition paper set
```

The 2x2 transience table (Table 3), the within-HA proportional novelty and
uptake quoted in the text, and the AA model-family check on scorer
familiarity, all computed from `novelty_scores.csv`.

Quantities, per turn $T_t$ followed by the partner's turn $F_{t+1}$:

- $S_{\text{base}}(T) = \bar{s}(T_t \mid \text{BOS})$, recovered as
  `surprise_raw - surprise` (the contextual score minus novelty)
- $\text{Novelty} = \bar{s}(T_t \mid C_t) - S_{\text{base}}(T)$, the `surprise` column
- $S(F) = \bar{s}(F_{t+1} \mid C_t)$, the `transience_baseline` column
- $\text{Transience} = \bar{s}(F_{t+1} \mid C_t, T_t) - S(F)$
- proportional context gain $= -\text{Novelty} / S_{\text{base}}(T)$
- uptake $= -\text{Transience} / S(F)$

Every percentage is a story-clustered mean: the per-turn ratio is averaged
within each story, then across stories.

```{r load}
df_raw <- load_novelty_data(CONDITIONS) %>%
  filter(!conversation_id %in% EXCLUDED_CONVERSATIONS)

# One row per turn, keeping which model wrote it where that is known.
slot_rows <- function(df) {
  bind_rows(lapply(c("author_1", "author_2"), function(slot) {
    other <- if (slot == "author_1") "author_2" else "author_1"
    # Cross-model stories: the model follows the starter role, since slots are
    # counterbalanced. Same-model stories have a single model_id.
    col_or_na <- function(name) if (name %in% names(df)) as.character(df[[name]]) else rep(NA_character_, nrow(df))
    by_role <- if_else(col_or_na("starter_side") == slot, col_or_na("model_starter"), col_or_na("model_responder"))
    model <- coalesce(by_role, col_or_na("model_id"))
    tibble(
      conversation_id = df$conversation_id,
      condition = df$condition,
      turn = df$turn,
      speaker = df[[paste0(slot, "_type")]],
      partner = df[[paste0(other, "_type")]],
      model = model,
      novelty = df[[paste0(slot, "_surprise")]],
      contextual = df[[paste0(slot, "_surprise_raw")]],
      transience = df[[paste0(slot, "_transience")]],
      base_F = df[[paste0(slot, "_transience_baseline")]]
    )
  })) %>%
    mutate(
      base_T = contextual - novelty,
      context_gain = -novelty / base_T,
      uptake = -transience / base_F,
      pairing = case_when(
        condition == "ai-ai-cross" ~ "AI -> AI (cross-model)",
        speaker == "human" & partner == "ai" ~ "Human -> AI",
        speaker == "ai" & partner == "human" ~ "AI -> Human",
        speaker == "human" & partner == "human" ~ "Human -> Human",
        speaker == "ai" & partner == "ai" ~ "AI -> AI"
      )
    )
}

turns <- slot_rows(df_raw)

story_mean <- function(df, ..., cols) {
  df %>%
    group_by(..., conversation_id) %>%
    summarise(across(all_of(cols), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
    group_by(...) %>%
    summarise(n_stories = n(), across(all_of(cols), ~ mean(.x, na.rm = TRUE)), .groups = "drop")
}
```

# Transience 2x2 (paper Table 3)

Speaker type crossed with partner type, turns after the first ($t > 1$).
Transience is measured on the partner's following turn, so who responds
constrains the score; comparing within a row of the 2x2 holds the speaker
fixed and varies the partner.

```{r transience-2x2}
pairing_order <- c("Human -> AI", "AI -> Human", "Human -> Human", "AI -> AI", "AI -> AI (cross-model)")

t22 <- turns %>%
  filter(turn > 1, !is.na(transience), !is.na(base_F)) %>%
  story_mean(pairing, cols = c("base_F", "transience", "uptake")) %>%
  mutate(pairing = factor(pairing, levels = pairing_order)) %>%
  arrange(pairing)

t22 %>%
  mutate(uptake = 100 * uptake) %>%
  knitr::kable(
    digits = 2,
    col.names = c("Pairing", "Stories", "Base S(F)", "Transience", "Uptake (%)"),
    caption = "Transience 2x2: speaker -> partner, story-clustered means"
  )

up <- setNames(100 * t22$uptake, as.character(t22$pairing))
```

Derived in-text contrasts:

- human turns, LLM partner vs human partner:
  `r sprintf("%+.1f", up[["Human -> AI"]] - up[["Human -> Human"]])` percentage points of uptake
- LLM turns, human partner vs LLM partner:
  `r sprintf("%+.1f", up[["AI -> Human"]] - up[["AI -> AI"]])` percentage points of uptake
- turns followed by a human writer:
  `r sprintf("%.1f%%-%.1f%%", min(up[["AI -> Human"]], up[["Human -> Human"]]), max(up[["AI -> Human"]], up[["Human -> Human"]]))`;
  turns followed by an LLM:
  `r sprintf("%.1f%%-%.1f%%", min(up[["Human -> AI"]], up[["AI -> AI"]]), max(up[["Human -> AI"]], up[["AI -> AI"]]))`

The AI -> AI uptake here is the story-clustered mean of per-turn ratios, like
the other three cells. The paper's 38.4% equals the ratio of the two means
beside it (`r sprintf("%.2f / %.2f = %.1f%%", -t22$transience[t22$pairing == "AI -> AI"], t22$base_F[t22$pairing == "AI -> AI"], -100 * t22$transience[t22$pairing == "AI -> AI"] / t22$base_F[t22$pairing == "AI -> AI"])`),
so that one cell appears to have been computed differently.

# Within Human-AI: how much of each turn the context explains

```{r within-ha}
turns %>%
  filter(condition == "human-ai", turn > 1) %>%
  story_mean(speaker, cols = c("novelty", "context_gain", "uptake")) %>%
  mutate(across(c(context_gain, uptake), ~ 100 * .x)) %>%
  knitr::kable(
    digits = c(0, 0, 2, 1, 1),
    col.names = c("Speaker", "Stories", "Novelty", "Context gain (%)", "Uptake (%)"),
    caption = "HA: share of turn information explained by prior context, and uptake by the partner"
  )
```

# Scorer familiarity: AA stratified by model family

Does the scorer (`google/gemma-4-31B`) simply find LLM text easy to predict?
If so, base surprisal would be low for every generator and the context gain
would track base surprisal. Stratifying by generator separates the two. This
check uses all turns, including the first, as in the paper.

```{r family-check}
family_rows <- turns %>%
  filter(!is.na(novelty), !is.na(base_T)) %>%
  mutate(group = case_when(
    speaker == "human" ~ paste0("Human (", condition, ")"),
    condition %in% c("ai-ai", "ai-ai-cross") ~ paste0(model, if_else(condition == "ai-ai-cross", " [cross]", "")),
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(group))

family_rows %>%
  story_mean(group, cols = c("base_T", "context_gain")) %>%
  mutate(context_gain = 100 * context_gain) %>%
  arrange(desc(base_T)) %>%
  knitr::kable(
    digits = 2,
    col.names = c("Writer", "Stories", "Base surprisal (bits/token)", "Context gain (%)"),
    caption = "Unconditional surprisal and proportional context gain by writer"
  )
```

In the cross-model rows, a story counts once for each model that wrote in
it, and the model is attributed per turn from `starter_side`, since the
author columns are counterbalanced.

The other robustness checks (shuffled context, lexical baselines, surface
PERMANOVA) are in `robustness_checks.Rmd`.
